\documentclass[10pt]{article}
\usepackage{titling}

\title{Computer Architecture}
\date{DATE PLACEHOLDER}
\author{Ingyu Bahng}

\input{../../presets/computer_science/preamble}

\begin{document}

\input{../../presets/_general/startup}

\section{Fundamental Units}
Bits and bytes.
Binary and hexadecimal number systems.
Logic gates (AND, OR, NOT, XOR).
Boolean algebra.

\section{CPU (Central Processing Unit)}
CPU components (ALU, control unit, registers, cache).
The fetch-decode-execute cycle.
Clock speed and performance.
Cores and threads.
Pipeline and superscalar execution.
Cache hierarchy (L1, L2, L3).

\section{Instruction Set Architecture (ISA)}
What an ISA is --- the contract between hardware and software.
x86/x86\_64 --- Intel/AMD, CISC architecture, dominant in desktops/servers.
ARM --- RISC architecture, dominant in mobile/embedded, Apple Silicon.
RISC-V --- open source ISA, emerging.
CISC vs RISC distinction.

\section{Memory}
Memory hierarchy (registers, cache, RAM, storage).
RAM types (DRAM, SRAM).
Virtual memory and paging.
Memory addressing.

\section{Storage}
HDD vs SSD.
NVMe vs SATA.
File systems (ext4, APFS, NTFS).

\section{Motherboard and Bus Systems}
Motherboard components.
PCIe, USB, SATA buses.
Northbridge/Southbridge (chipset).

\section{GPU}
CPU vs GPU architecture.
Parallel vs sequential processing.
CUDA and compute.

\section{Input/Output}
I/O controllers.
Interrupts and polling.
DMA (Direct Memory Access).

\section{Power}
TDP (Thermal Design Power).
Power delivery.
Thermal management.
  
\end{document}

